Type Classes for Mathematics in Type Theory

### 2010-07-01 jar.tex
Improvements/corrections suggested by James.

The base of our development consists of type classes representing a standard algebraic hierarchy, as well as portions of category theory and universal algebra. On this foundation we build a set of mathematically sound abstract interfaces for different kinds of numbers, succinctly expressed using categorical language and universal algebra constructions. Strategic use of type classes lets us support these high-level theory-friendly definitions while still enabling efficient implementations unhindered by gratuitous indirection or conversion or projection.

Algebra flourishes by the interplay between syntax and semantics. The Prolog-like abilities of type class instance resolution allow us to conveniently define a quote function, thus facilitating the use of reflective techniques.
\end{abstract}

@@

Propositions in Coq are types~\cite{ITT,CMCP}, which themselves have types called \emph{sorts}. Coq features a distinguished sort called \lstinline|Prop| that one may choose to use as the sort for types representing propositions. The distinguishing feature of the \lstinline|Prop| sort is that terms of non-\lstinline|Prop| type may not depend on the values of inhabitants of \lstinline|Prop| types (that is, proof terms).
% As originally conceived~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
% them. Often we do \emph{not} want functions to depend on proofs. In Coq, this is the paradigmatic way of working. We are forbidden to extract information from proofs of \lstinline|Prop|s. Especially with the aid of the Program machinery, this allows us to flexibly work with simply typed program and their Hoare style correctness proofs.
This regime of discrimination establishes a weak form of proof irrelevance, in that changing a proof can never affect the result of value computations. On a very practical level, this lets Coq safely erase all \lstinline|Prop| components when extracting certified programs to OCaml or Haskell.
A stronger notion of proof irrelevance in which any two inhabitants of a \lstinline|Prop| type are made \emph{convertible} is planned for a future Coq version.
% todo: process?:
%One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
%construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. In Coq,
%the [ ]-types are isomorphic to the type of sort \lstinline|Prop|. However,
%there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
%theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
%irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
%of Propositions as types seems necessary.
% Making Harrop formulas proof irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. An orthogonal and further reaching proposal is to replace the Coq type theory with the Calculus of implicit constructions with implicit Σ-types~\cite{miquel2001implicit,barras2008implicit,Bernardo}.

@@

The `native' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor
\begin{lstlisting}
 eq_refl : Π (T: Type) (x: T), x ≡ x,
\end{lstlisting}
where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|. Here we diverge from Coq tradition and reserve \lstinline|=| for setoid equality, as this is the equality we will be working with most of the time. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions. We mention this explicitly only because rewriting \emph{does} give rise to conditions when we depart from Leibniz equality and introduce equivalence relations representing abstractions over common representations of the same conceptual object. For instance, consider formal fractions of integers as a representation of rationals. Rewriting a subterm using such an equality is permitted only if the subterm is an argument of a function that has been proven to \emph{respect} the equality. Such a function is called \emph{proper}, and propriety must be proved for each function in whose arguments we wish to enable rewriting.

@@

The algebraic hierarchy of the ssreflect~\cite{Packed} libraries uses an interesting (if fairly restrictive) alternative approach. It requires unicity of representation for all objects so that Leibniz equality is always applicable.

@@

Algebraic structures are expressed in terms of a number of carriers, a number of operations and relations on these carriers, and a number of laws that the operations and relations satisfy. In a system like Coq, we have different options when it comes to representing the grouping of these components. On one end of the spectrum, we can simply define the (conjunction of) laws as an $n$-ary predicate over $n$ components, forgoing explicit grouping altogether. For instance, for the mundane example of a reflexive relation, we could use:

@@

More elaborate structures, too, can be expressed as predicates (expressing laws) over a number of carriers, relations, and operations. While optimally flexible in principle, in practice \emph{naive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest. And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time. Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience. Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.

@@

In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and Ssreflect~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind $+$ and $*$ notations to the record field projections for the addition and multiplication operations, respectively.

@@

We will tackle the problems associated with the structure predicates approach one by one, starting with those encountered during theorem \emph{application}. Suppose we have defined \lstinline|SemiGroup| as a structure predicate as follows\footnote{Note that defining \lstinline|SemiGroup| as a record instead of a straight conjunction does not make it any less of a predicate. The record form is simply more convenient in that it immediately gives us named projections for laws and substructures.}:
\begin{lstlisting}
  Record SemiGroup (G: Type) (e: relation G) (op: G → G → G): Prop :=
    { sg_setoid: Equivalence e
    ; sg_ass: Associative op
    ; sg_proper: Proper (e ==> e ==> e) op }.
\end{lstlisting}
Then by (1) making \lstinline|SemiGroup| a \emph{class} (by replacing the \lstinline|Record| keyword with the \lstinline|Class| keyword), (2) marking its proofs as \emph{instances} (again by replacing a single keyword), and (3) marking the \lstinline|SemiGroup| parameter of semigroup theorems as implicit (by using curly instead of round brackets), we no longer have to pass \lstinline|SemiGroup| proofs around manually ourselves, letting instance resolution do it for us instead. Because instance resolution is part of the unifier, this also works when the statement of the theorem we wish to apply only mentions some of the components (which admittedly doesn't make much sense for semigroups).

@@

The declarations of the two inherited \lstinline|CommutativeMonoid| structures nicely illustrate how predicate classes naturally support not just multiple inheritance, but \emph{overlapping} multiple inheritance, where the inherited structures may share components (in this case carrier and equivalence relation). The carrier \lstinline|A|, being an explicit argument, is specified as normal. The equivalence relation, being an implicit argument of class type, is resolved automatically to \lstinline|e| and passed. The binary operation and constant would normally be automatically resolved and passed as well, but we override the inference mechanism locally using Coq's existing named argument facility (which is only syntactic sugar of the most superficial kind) in order to specify the correct pairings. 
Again, contrast this with type system extensions such as Matita's manifest records, which are required to make this work when the records bundle components such as \lstinline|op| and \lstinline{unit} as \emph{fields} instead of parameters.

@@

To characterize structures such as semirings and rings, we need equational theories, consisting of a signature together with laws (represented by a predicate over equality entailments):
\begin{lstlisting}
  Record EquationalTheory :=
    { et_sig:> Signature
    ; et_laws:> EqEntailment et_sig → Prop }.
\end{lstlisting}
An \lstinline|EqEntailment| consists of premises and a conclusion represented by an inductively defined statement grammar, which in turn uses an inductively defined term grammar. A detailed discussion of these definitions and the theory developed for them is beyond the scope of this paper.

We now introduce a predicate class designating algebras that satisfy the laws of an equational theory:
\begin{lstlisting}
  Class InVariety
    (et: EquationalTheory) (carriers: sorts et → Type)
    {e: Π a, Equiv (carriers a)} `{!AlgebraOps et carriers}: Prop :=
    { variety_algebra:> Algebra et carriers
    ; variety_laws: Π s, et_laws et s → (Π $\hspace{-1.5mm}$vars, eval_stmt et vars s) }.
\end{lstlisting}
  % todo: why do i need the manual \hspace!?

What remains is to show that carriers+\lstinline|Equivs|+\lstinline|AlgebraOps| satisfying \lstinline|InVariety| for a given \lstinline|EquationalTheory| do indeed form a \lstinline|Category| (the `variety'). Since we need a type for the objects in the \lstinline|Category|, at this point we have no choice but to bundle components and proof together in a record:
\begin{lstlisting}
  Variable et: EquationalTheory.

@@

  Instance: ClosedSubset image.
\end{lstlisting}
\lstinline|sigT| represents a \lstinline|Type|-sorted existential. \lstinline|ClosedSubset| is defined elsewhere as:
\begin{lstlisting}
  Context `{Algebra σ$$ A} (P: Π s, A s → Type).
  Class ClosedSubset: Type :=
    { subset_proper: Π s x x', x = x' → iffT (P s x) (P s x')
    ; subset_closed: Π o, op_closed (algebra_op o) }.
\end{lstlisting}
\lstinline|op_closed| is defined by recursion over the symbolic operation types.

The reason we define \lstinline|image| and \lstinline|ClosedSubset| in \lstinline|Type| rather than \lstinline|Prop| is that since the final goal of the proof is to establish an isomorphism in the category of σ-algebras (where arrows are algebra homomorphisms), we will eventually need to map elements in the subalgebra defined by \lstinline|image| back to their pre-image in \lstinline|A|.

@@

Taking these conversion functions for granted, we will also need a ``down-to-earth'' representation of the initiality arrows if we are to give a \lstinline|SemiRing|-based definition of the interface for natural numbers. Once again, we introduce an operational type class to represent this particular component:
\begin{lstlisting}
  Class NaturalsToSemiRing (A: Type) :=
    naturals_to_semiring: Π B `{RingMult B} `{RingPlus B} `{RingOne B}
      `{RingZero B}, A → B.
\end{lstlisting}
To use this with \lstinline|Initial|, we define an additional conversion instance that takes a \lstinline|NaturalsToSemiRing| along with a proof showing that it yields \mbox{\lstinline|SemiRing_Morphism|s}, and builds an \lstinline|InitialArrow| instance out of it. This conversion instance in turn invokes another conversion function that translates concrete \lstinline|SemiRing_Morphism| proofs into univeral algebra \lstinline|Homomorphism|s instantiated with the semiring signature, which make up the arrows in the category.

@@

Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support this quoting. Unification hints~\cite{Hints}, a very general way of facilitating user-defined extensions to term and type inference, can be used to semi-automatically build quote functions without dropping to a meta-level.\footnote{Gonthier provides similar functionality by ingeniously using canonical structures.} This feature is absent from Coq, but fortunately type classes also allow us to do this, as we will now show.

@@

For the present example, we define an ad-hoc term language for monoids
\begin{lstlisting}
  Inductive Expr (V: Type) := Mult (a b: Expr V) | One | Var (v: V).
\end{lstlisting}
The expression type is parameterized over the set of variable indices. Below we use an implicitly defined heap of such variables. Hence, we diverge from~\cite{Hints}, which uses \lstinline|nat| for variables, thereby introducing bound problems and a need for dummy variables.

Suppose now that we want to quote \lstinline|nat| expressions built from 1 and multiplication. To describe the relation we want the symbolic expression to have to the original expression, we first define how symbolic expressions evaluate to values (given a variable assignment):
%An expression is only meaningful in the context of a variable assignment:\footnote{Categorically: \lstinline|nat| is an \lstinline|Expr|-algebra and \lstinline|eval| is \lstinline|map vars| composed with the algebra map.}
\begin{lstlisting}
  Definition Value := nat.
  Definition Vars V := V → Value.

  Fixpoint eval {V} (vs: Vars V) (e: Expr V): Value :=
    match e with
    | One => 1
    | Mult a b => eval vs a * eval vs b
    | Var v => vs v
    end.
\end{lstlisting}

@@

We can now state our goal: given an expression of type \lstinline|nat|, we seek to construct an \lstinline|Expr V| for some appropriate \lstinline|V| along with a variable assignment, such that evaluation of the latter yields the former. Because we will be doing this incrementally, we introduce a few simple variable ``heap combinators'':
\begin{lstlisting}
  Definition novars: Vars False := False_rect _.
  Definition singlevar (x: Value): Vars unit := λ _ => x.
  Definition merge {A B} (a: Vars A) (b: Vars B): Vars (A+B) :=
    λ i => match i with inl j => a j | inr j => b j end.
\end{lstlisting}
These last two combinators are the `constructors' of an implicitly defined subset of Gallina terms, representing heaps, for which we will implement syntactic lookup with type classes in a moment. The heap can also be defined explicitly, with no essential change in the code.

With these, we can define the primary ingredient, the \lstinline|Quote| class:
\begin{lstlisting}
  Class Quote {V} (l: Vars V) (n: Value) {V'} (r: Vars V'): Type :=
    { quote: Expr (V + V')
    ; eval_quote: eval (merge l r) quote = n }.
\end{lstlisting}
We can think of \lstinline|Quote| as the type for a family of Prolog-like syntax-directed resolution functions, which will take as input \lstinline|V| and \lstinline|l| representing previously encountered holes (opaque subexpressions that could not be destructured further) and their values, along with a concrete term \lstinline|n| to be quoted. Their `output' will consist not only of the fields in the class, but also of \lstinline|V'| and \lstinline|r| representing additional holes and their values. Hence, a type class constraint of the form \lstinline|Quote x y z| should be read as ``quoting \lstinline|y| with existing heap \lstinline|x| generates new heap \lstinline|z|''.

The \lstinline|Quote| instance for 1 illustrates the basic idea:
\begin{lstlisting}
  Instance quote_one V (v: Vars V): Quote v 1 novars := { quote := One }.
\end{lstlisting}
The expression `1' can be quoted in any context \lstinline|(V, v)|, introduces no new variables, and the symbolic term representing it is just \lstinline|One|. The \lstinline|eval_quote| field is turned into a trivial proof obligation.

The \lstinline|Quote| instance for multiplication is a little more subtle, but really only does a bit of heap juggling:
\begin{lstlisting}
  Instance quote_mult V (v: Vars V) n V' (v': Vars V') m V'' (v'': Vars V'')
    `{!Quote v n v'} `{!Quote (merge v v') m v''}:
      Quote v (n * m) (merge v' v'') :=
      { quote :=
        Mult (map_var shift (quote n)) (map_var sum_assoc (quote m)) }.
\end{lstlisting}

These two instances specify how 1 and multiplications are to be quoted, but what about other expressions? For these, we want to distinguish two kinds: expressions we have seen before, and those we have not. To make this distinction, we need to be able to look up expressions in variable heaps to see if they're already there. Importantly, we must do this not by comparing the values they evaluate to, but by actually browsing the term denoting the variable heap --- that is, a composition from \lstinline|novars|, \lstinline|singlevar|, and \lstinline|merge|. This, too, is a job for a type class:
\begin{lstlisting}
  Class Lookup {A} (x: Value) (v: Vars A) := { key: A; key_correct: v key = x }.
\end{lstlisting}
Our first \lstinline|Lookup| instance states that \lstinline|x| can be looked up in \lstinline|singlevar x|:
\begin{lstlisting}
  Instance singlevar_lookup (x: Value): Lookup x (singlevar x) := { key := tt }.
\end{lstlisting}
Finally, if an expression can be looked up in a pack, then it can also be looked up when that pack is merged with another pack:
\begin{lstlisting}
  Context (x: Value) {A B} (va: Vars A) (vb: Vars B).

@@

With \lstinline|Lookup|, we can now define a \lstinline|Quote| instance for previously encountered expressions:
\begin{lstlisting}
  Instance quote_old_var V (v: Vars V) x {Lookup x v}:
    Quote v x novars | 8 := { quote := Var (inl (key x v)) }.
\end{lstlisting}

If none of the \lstinline|Quote| instances defined so far apply, the term in question is a newly encountered hole. For this case we define a catch-all instance with a low priority, which yields a singleton heap containing the expression:
\begin{lstlisting}
  Instance quote_new_var V (v: Vars V) x: Quote v x (singlevar x) | 9
    := { quote := Var (inr tt) }.
\end{lstlisting}
And with that, we can start quoting:
\begin{lstlisting}
  Goal Π x y (P: Value → Prop), P ((x * y) * (x * 1)).
    intros.
    rewrite <- eval_quote.
\end{lstlisting}
The \lstinline|rewrite| rewrites the goal to (something that reduces to):
\begin{lstlisting}
  P (eval
     (merge novars
        (merge (merge (singlevar x) (singlevar y)) (merge novars novars)))
     (Mult (Mult (Var (inr (inl (inl ())))) (Var (inr (inl (inr ())))))
        (Mult (Var (inr (inl (inl ())))) One)))
\end{lstlisting}

Using a small additional utility lemma we can also easily quote equalities into \lstinline|eval p e = eval p e'| form with a shared variable pack (so that identical expressions on both sides of the equation don't end up as different variables).

>>>>

The base of our development consists of type classes representing a standard algebraic hierarchy, as well as portions of category theory and universal algebra. On this foundation we build a set of mathematically sound abstract interfaces for different kinds of numbers, succinctly expressed using categorical language and universal algebra constructions. Strategic use of type classes lets us support these high-level theory-friendly definitions while still enabling efficient implementations unhindered by gratuitous indirection, conversion or projection.

Algebra thrives on the interplay between syntax and semantics. The Prolog-like abilities of type class instance resolution allow us to conveniently define a quote function, thus facilitating the use of reflective techniques.
\end{abstract}

@@

Propositions in Coq are types~\cite{ITT,CMCP}, which themselves have types called \emph{sorts}. Coq features a distinguished sort called \lstinline|Prop| that one may choose to use as the sort for types representing propositions. The distinguishing feature of the \lstinline|Prop| sort is that terms of non-\lstinline|Prop| type may not depend on the values of inhabitants of \lstinline|Prop| types (that is, proof terms).
% As originally conceived~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
% them. Often we do \emph{not} want functions to depend on proofs. In Coq, this is the paradigmatic way of working. We are forbidden to extract information from proofs of \lstinline|Prop|s. Especially with the aid of the Program machinery, this allows us to flexibly work with simply typed program and their Hoare style correctness proofs.
This regime of discrimination establishes a weak form of proof irrelevance, in that changing a proof can never affect the result of value computations. On a very practical level, this lets Coq safely erase all \lstinline|Prop| components when extracting certified programs to OCaml or Haskell.

% A stronger notion of proof irrelevance in which any two inhabitants of a \lstinline|Prop| type are made \emph{convertible} is planned for a future Coq version.
  % commented out because not really relevant. only re-add with explanation of relevance.

% todo: process?:
%One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
%construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. In Coq,
%the [ ]-types are isomorphic to the type of sort \lstinline|Prop|. However,
%there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
%theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
%irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
%of Propositions as types seems necessary.
% Making Harrop formulas proof irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. An orthogonal and further reaching proposal is to replace the Coq type theory with the Calculus of implicit constructions with implicit Σ-types~\cite{miquel2001implicit,barras2008implicit,Bernardo}.

@@

The `native' notion of equality in Coq is that of terms being convertible, naturally reified as a proposition by the inductive type family \lstinline|eq| with single constructor \lstinline|eq_refl|:
\begin{lstlisting}
  eq_refl : Π (T: Type) (x: T), x ≡ x,
\end{lstlisting}
where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|. Here we diverge from Coq tradition and reserve \lstinline|=| for setoid equality, as this is the equality we will be working with most of the time. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions. We mention this explicitly only because rewriting \emph{does} give rise to conditions when we depart from Leibniz equality and introduce equivalence relations representing abstractions over common representations of the same conceptual object. For instance, consider formal fractions of integers as a representation of rationals. Rewriting a subterm using such an equality is permitted only if the subterm is an argument of a function that has been proven to \emph{respect} the equality. Such a function is called \emph{proper}, and propriety must be proved for each function in whose arguments we wish to enable rewriting.

@@

The algebraic hierarchy of the \textsc{Ssreflect} libraries~\cite{Packed} uses an interesting (if fairly restrictive) alternative approach. It requires unicity of representation for all objects so that Leibniz equality is always applicable.

@@

Algebraic structures are expressed in terms of a number of carrier sets, a number of operations and relations on these carriers, and a number of laws that the operations and relations satisfy. In a system like Coq, we have different options when it comes to representing the grouping of these components. On one end of the spectrum, we can simply define the (conjunction of) laws as an $n$-ary predicate over $n$ components, forgoing explicit grouping altogether. For instance, for the mundane example of a reflexive relation, we could use:

@@

More elaborate structures, too, can be expressed as predicates (expressing laws) over a number of carriers, relations, and operations. While optimally flexible in principle, in practice \emph{na\"ive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest. And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time. Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience. Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.

@@

In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and \textsc{Ssreflect}~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind $+$ and $*$ notations to the record field projections for the addition and multiplication operations, respectively.

@@

We will tackle the problems associated with the use of structure predicates one by one, starting with those encountered during theorem \emph{application}. Suppose we have defined \lstinline|SemiGroup| as a structure predicate as follows\footnote{Note that defining \lstinline|SemiGroup| as a record instead of a straight conjunction does not make it any less of a predicate. The record form is simply more convenient in that it immediately gives us named projections for laws and substructures.}:
\begin{lstlisting}
  Record SemiGroup (G: Type) (e: relation G) (op: G → G → G): Prop :=
    { sg_setoid: Equivalence e
    ; sg_ass: Associative op
    ; sg_proper: Proper (e ==> e ==> e) op }.
\end{lstlisting}
Then by (1) making \lstinline|SemiGroup| a \emph{class} (by replacing the \lstinline|Record| keyword with the \lstinline|Class| keyword), (2) marking its proofs as \emph{instances} (again by replacing a single keyword), and (3) marking the \lstinline|SemiGroup| parameter of semigroup theorems as implicit (by using curly instead of round brackets), we no longer have to pass \lstinline|SemiGroup| proofs around manually ourselves, letting instance resolution do it for us instead. Because instance resolution is part of the unifier, this also works when the statement of the theorem we wish to apply only mentions some of the components (which admittedly doesn't make much sense for semigroups).

@@

The declarations of the two inherited \lstinline|CommutativeMonoid| structures in \lstinline|SemiRing| nicely illustrate how predicate classes naturally support not just multiple inheritance, but \emph{overlapping} multiple inheritance, where the inherited structures may share components (in this case carrier and equivalence relation). The carrier \lstinline|A|, being an explicit argument, is specified as normal. The equivalence relation, being an implicit argument of class type, is resolved automatically to \lstinline|e|. The binary operation and constant would normally be automatically resolved as well, but we override the inference mechanism locally using Coq's existing named argument facility (which is only syntactic sugar of the most superficial kind) in order to explicitly pair multiplication with 1 for the first \lstinline|CommutativeMonoid| substructure, and addition with 0 for the second \lstinline|CommutativeMonoid| substructure. Again, contrast this with type system extensions such as Matita's manifest records, which are required to make this work when the records bundle components such as \lstinline|op| and \lstinline{unit} as \emph{fields} instead of parameters.

@@

To characterize structures such as semirings and rings, we need equational theories, consisting of a signature together with laws (represented by a predicate over equality entailments):
\begin{lstlisting}
  Record EquationalTheory :=
    { eqt_sig:> Signature
    ; eqt_laws:> EqEntailment eqt_sig → Prop }.
\end{lstlisting}
An \lstinline|EqEntailment| consists of premises and a conclusion represented by an inductively defined statement grammar, which in turn uses an inductively defined term grammar. A detailed discussion of these definitions and the theory developed for them is beyond the scope of this paper.

We now introduce a predicate class designating algebras that satisfy the laws of an equational theory:
\begin{lstlisting}
  Class InVariety
    (et: EquationalTheory) (carriers: sorts et → Type)
    {e: Π a, Equiv (carriers a)} `{!AlgebraOps et carriers}: Prop :=
    { variety_algebra:> Algebra et carriers
    ; variety_laws: Π s, eqt_laws et s → (Π $\hspace{-1.5mm}$vars, eval_stmt et vars s) }.
\end{lstlisting}
  % todo: why do i need the manual \hspace!?

What remains is to show that carrier sets together with \lstinline|Equiv|s and \lstinline|AlgebraOps| satisfying \lstinline|InVariety| for a given \lstinline|EquationalTheory| do indeed form a \lstinline|Category| (the `variety'). Since we need a type for the objects in the \lstinline|Category|, at this point we have no choice but to bundle components and proof together in a record:
\begin{lstlisting}
  Variable et: EquationalTheory.

@@

  Instance: ClosedSubset image.
\end{lstlisting}
The \lstinline|sigT| type constructor is a \lstinline|Type|-sorted existential quantifier. \lstinline|ClosedSubset| is defined elsewhere as:
\begin{lstlisting}
  Context `{Algebra σ$$ A} (P: Π s, A s → Type).
  Class ClosedSubset: Type :=
    { subset_proper: Π s x x', x = x' → iffT (P s x) (P s x')
    ; subset_closed: Π o, op_closed (algebra_op o) }.
\end{lstlisting}
Here, \lstinline|op_closed| is defined by recursion over the symbolic operation types.

The reason we define \lstinline|image| and \lstinline|ClosedSubset| in \lstinline|Type| rather than in \lstinline|Prop| is that since the final goal of the proof is to establish an isomorphism in the category of σ-algebras (where arrows are algebra homomorphisms), we will eventually need to map elements in the subalgebra defined by \lstinline|image| back to their pre-image in \lstinline|A|.

@@

Taking these conversion functions for granted, we will also need a ``down-to-earth'' representation of the initiality arrows if we are to give a \lstinline|SemiRing|-based definition of the interface for natural numbers. Once again, we introduce an operational type class to represent this particular component:
\begin{lstlisting}
  Class NaturalsToSemiRing (A: Type) :=
    naturals_to_semiring: Π B `{RingMult B} `{RingPlus B} `{RingOne B}
      `{RingZero B}, A → B.
\end{lstlisting}
The instance for \lstinline|nat| is defined as follows:
\begin{lstlisting}
Instance nat_to_semiring: NaturalsToSemiRing nat :=
  λ _ _ _ _ _ => fix f (n: nat) := match n with 0 => 0 | S m => f m + 1 end.
\end{lstlisting}

To use \lstinline|NaturalsToSemiRing| with \lstinline|Initial|, we define an additional conversion instance that takes a \lstinline|NaturalsToSemiRing| along with a proof showing that it yields \mbox{\lstinline|SemiRing_Morphism|s}, and builds an \lstinline|InitialArrow| instance out of it. This conversion instance in turn invokes another conversion function that translates concrete \lstinline|SemiRing_Morphism| proofs into univeral algebra \lstinline|Homomorphism|s instantiated with the semiring signature, which make up the arrows in the category.

@@

Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support such \emph{quoting}. Unification hints~\cite{Hints}, a very general way of facilitating user-defined extensions to term and type inference, can be used to semi-automatically build quote functions without dropping to a meta-level.\footnote{Gonthier provides similar functionality by ingeniously using canonical structures.} This feature is absent from Coq, but fortunately type classes also allow us to do this, as we will now show.

@@

For the present example, we define an ad-hoc term language for monoids
\begin{lstlisting}
  Inductive Expr (V: Type) := Mult (a b: Expr V) | One | Var (v: V).
\end{lstlisting}
The expression type is parameterized over the set of variable indices. Below we use an implicitly defined heap of such variables. Hence, we diverge from~\cite{Hints}, which uses \lstinline|nat| for variable indices, thereby introducing a need for dummy variables for out-of-bounds indices.

Suppose now that we want to quote \lstinline|nat| expressions built from 1 and multiplication. To describe the relation we want the symbolic expression to have to the original expression, we first define how symbolic expressions evaluate to values (given a variable assignment):
%An expression is only meaningful in the context of a variable assignment:\footnote{Categorically: \lstinline|nat| is an \lstinline|Expr|-algebra and \lstinline|eval| is \lstinline|map vars| composed with the algebra map.}
\begin{lstlisting}
  Definition Value := nat.
  Definition Env V := V → Value.

  Fixpoint eval {V} (vs: Env V) (e: Expr V): Value :=
    match e with
    | One => 1
    | Mult a b => eval vs a * eval vs b
    | Var v => vs v
    end.
\end{lstlisting}

@@

We can now state our goal: given an expression of type \lstinline|nat|, we seek to construct an \lstinline|Expr V| for some appropriate \lstinline|V| along with a variable assignment, such that evaluation of the latter yields the former. Because we will be doing this incrementally, we introduce a few simple variable ``heap combinators'':
\begin{lstlisting}
  Definition novars: Env False := False_rect _.
  Definition singlevar (x: Value): Env unit := λ _ => x.
  Definition merge {A B} (a: Env A) (b: Env B): Env (A+B) :=
    λ i => match i with inl j => a j | inr j => b j end.
\end{lstlisting}
These last two combinators are the `constructors' of an implicitly defined subset of Gallina terms, representing heaps, for which we will implement syntactic lookup with type classes in a moment. The heap can also be defined explicitly, with no essential change in the code.

With these, we can define the primary ingredient, the \lstinline|Quote| class:
\begin{lstlisting}
  Class Quote {V} (l: Env V) (n: Value) {V'} (r: Env V'): Type :=
    { quote: Expr (V + V')
    ; eval_quote: eval (merge l r) quote = n }.
\end{lstlisting}
We can think of \lstinline|Quote| as the type for a family of Prolog-like syntax-directed resolution functions, which will take as input \lstinline|V| and \lstinline|l| representing previously encountered holes (opaque subexpressions that could not be destructured further) and their values, along with a concrete term \lstinline|n| to be quoted. Their `output' will consist not only of the fields in the class, but also of \lstinline|V'| and \lstinline|r| representing additional holes and their values. Hence, a type class constraint of the form \lstinline|Quote x y z| should be read as ``quoting \lstinline|y| with existing heap \lstinline|x| generates new heap \lstinline|z|''.

The \lstinline|Quote| instance for 1 illustrates the basic idea:
\begin{lstlisting}
  Instance quote_one V (v: Env V): Quote v 1 novars := { quote := One }.
\end{lstlisting}
The expression `1' can be quoted in any context \lstinline|(V, v)|, introduces no new variables, and the symbolic term representing it is just \lstinline|One|. The \lstinline|eval_quote| field is turned into a trivial proof obligation.

The \lstinline|Quote| instance for multiplication is a little more subtle, but really only does a bit of heap juggling:
\begin{lstlisting}
  Instance quote_mult V (v: Env V) n V' (v': Env V') m V'' (v'': Env V'')
    `{!Quote v n v'} `{!Quote (merge v v') m v''}:
      Quote v (n * m) (merge v' v'') :=
      { quote :=
        Mult (map_var shift (quote n)) (map_var sum_assoc (quote m)) }.
\end{lstlisting}

These two instances specify how 1 and multiplications are to be quoted, but what about other expressions? For these, we want to distinguish two kinds: expressions we have seen before, and those we have not. To make this distinction, we need to be able to look up expressions in variable heaps to see if they're already there. Importantly, we must do this not by comparing the values they evaluate to, but by actually browsing the term denoting the variable heap --- that is, a composition from \lstinline|novars|, \lstinline|singlevar|, and \lstinline|merge|. This, too, is a job for a type class:
\begin{lstlisting}
  Class Lookup {A} (x: Value) (v: Env A) := { key: A; key_correct: v key = x }.
\end{lstlisting}
Our first \lstinline|Lookup| instance states that \lstinline|x| can be looked up in \lstinline|singlevar x|:
\begin{lstlisting}
  Instance singlevar_lookup (x: Value): Lookup x (singlevar x) := { key := tt }.
\end{lstlisting}
Finally, if an expression can be looked up in a pack, then it can also be looked up when that pack is merged with another pack:
\begin{lstlisting}
  Context (x: Value) {A B} (va: Env A) (vb: Env B).

@@

With \lstinline|Lookup|, we can now define a \lstinline|Quote| instance for previously encountered expressions:
\begin{lstlisting}
  Instance quote_old_var V (v: Env V) x {Lookup x v}:
    Quote v x novars | 8 := { quote := Var (inl (key x v)) }.
\end{lstlisting}

If none of the \lstinline|Quote| instances defined so far apply, the term in question is a newly encountered hole. For this case we define a catch-all instance with a low priority, which yields a singleton heap containing the expression:
\begin{lstlisting}
  Instance quote_new_var V (v: Env V) x: Quote v x (singlevar x) | 9
    := { quote := Var (inr tt) }.
\end{lstlisting}
And with that, we can start quoting:
\begin{lstlisting}
  Goal Π x y (P: Value → Prop), P ((x * y) * (x * 1)).
    intros.
    rewrite <- eval_quote.
\end{lstlisting}
The \lstinline|rewrite| rewrites the goal to (something that reduces to):
\begin{lstlisting}
  P (eval
     (merge novars
        (merge (merge (singlevar x) (singlevar y)) (merge novars novars)))
     (Mult (Mult (Var (inr (inl (inl ())))) (Var (inr (inl (inr ())))))
        (Mult (Var (inr (inl (inl ())))) One)))
\end{lstlisting}
